A tutorial on Generalised Additive Mixed Effects Models for bilingualism research

### 2023-12-12 manuscript/manuscript.qmd
add vowel lit

\[LIT REVIEW\]

We have simulated phonetic data from three groups of bilinguals: simultaneous English/Spanish bilinguals, beginners late English/Spanish bilinguals and advanced late English/Spanish bilinguals. The data contains the Euclidean Distance (EuD) of three Spanish vowels /a, i, u/ from the vowel space centroid, taken from word-final non-stressed vowels, along 9 equidistant time points. The measure is a good proxy for vowel reduction phenomena, which is to be expected in late bilinguals.

>>>>

To motivate our next example, we will now consider the production of Spanish vowels in three groups of bilinguals: simultaneous/native English/Spanish bilinguals (henceforth `es`) and late learners of Spanish with low and high levels of proficiency, `es_0` and `es_1`, respectively. Spanish has five phonemic vowels, /i, e, a, o, u/, and is often described as being a syllable-timed language in which the spectral evelope is rather stable and stressed and unstressed syllables have approximately the same duration. Though there is evidence that unstressed vowels can centralize to a certain degree, the differences between stressed and unstressed vowels are believed to be imperceptible [@martinez_celdran_1984]. American English, on the other hand, has a larger vowel inventory and is often described as a stress-timed language in which the duration between stressed syllables is approximately the same. Vowel reduction can occur in unstressed syllables, which typically manifests via shortening and/or centralization, with unstressed vowels often reducing to schwa ([ə]). Accordingly, English speaking learners of Spanish face a substantial obstacle when it comes to producing and perceiving Spanish vowels [See @aldrich2014acquisition; @cobb2009pronunciacion; @cobb2015adult; @bland2016speech; @iruela1997adquisicion, among others]. To wit, they often display cross-linguistic influence by producing unstressed Spanish vowels with \[ə\], particularly in the case of /a/, e.g., "casa" (Eng. *house*) as \[ka.sə\], and diphthongizing vowels in final position [@cobb2015adult].

In the following example, we consider the production of Spanish vowels in the aforementioned groups of bilingual speakers. Interestingly, to our knowledge, all of the current research investigating the cross-linguistic influence of vowel reduction processes in learners of Spanish have utilized modelling strategies that rely upon formant measurements at a single time point, typically the mid-point of the vowel, rather than scrutinizing the entire trajectory of the spectral envelope. In our view, the dynamic nature of vowel formants poses an interesting use-case for modelling using non-linear methods, such as GAMs. To this end, we have simulated phonetic data from three groups of bilinguals: simultaneous English/Spanish bilinguals, beginners late English/Spanish bilinguals and advanced late English/Spanish bilinguals. The data contains the Euclidean Distance (EuD) of three Spanish vowels /a, i, u/ from the vowel space centroid, taken from word-final non-stressed vowels, along 9 equidistant time points. The measure is a good proxy for vowel reduction phenomena, which is to be expected in late bilinguals.
